NDNS multilevel LCA CW2CB2 or CW2CB3

### 2018-08-31 paper/Chapters/Chapter3.tex
paper modified accordingly

A series of traditional LCA of the responses to carbohydrate intake within the 7 time slots of the day was first examined. These initial analyses ignored the clustering of observation days within participants of the survey. \textbf{Table \ref{tab:mixmodels}} shows the latent class solutions for one to five classes (see rows under the Fixed effects model section). The BIC declines with increases in the number of day level classes increases. However, the improvement of BIC dropped to less than 1000 from 3 class to 4 class solutions (658.9) and from 4 class to 5 class solutions (361.7). Entropy index indicates that the 4 class model could explain about 51\% percent of the data, while \textit{p} values of Lo-Mendell-Rubun LRT suggest that the more classes we fit, the better model we will have until up to 6 classes (\textit{p} = 0.06 and is not shown in the table). For the sake of parsimony, we only extended the model with random effects building on 2 class, 3 class, and 4 class solutions.

@@

As expected, consumption of total energy (7539.98 kJ), total carbohydrate (222.79 g), total sugar (98.63 g), starch (124.07 g), and non-milk extrinsic sugar (59.45 g) were the highest among high percentage carbohydrate days (all \textit{p} < 0.001). On the other hand, the consumption of protein (73.29 g), total fat (67.24 g), and alcohol (13.80 g) were the highest in the regular meals days. Moreover, in the high percentage carbohydrate days, participants turned to consume the highest amount of fruit (107.40 g). There was no evidence of any difference in the consumption of yellow, red, or green vegetables across the three types of days (\textit{p} = 0.681).\vspace{-0.6cm}

@@

With three individual-level latent classes \textbf{(Figure \ref{fig:level2})}, a low-carbohydrate eaters class, a moderate-carbohydrate eaters class, and a high-carbohydrate eaters class emerge. 43.1\% participants were identified as high-carbohydrate eaters, in these individuals, about 50\% of the days (2 out of 4 days) of their dietary diary could be classified as having high percentage carbohydrate days. Nearly 1 out of 4 days of their dietary diary were either "regular meals day" or "low carbohydrate day". 28.1\% participants fell into the low carbohydrate eaters class on the left-hand side of \textbf{Figure \ref{fig:level2}}, their recordings of food intake showed that in more than 60\% of their days, they were having "regular meals" which was characterised as with the highest amount of fat and alcohol consumptions as already described in \textbf{Table \ref{tab:day-level-features}}. Moderate carbohydrate eaters have comparable proportions (42.0\% vs. 40.0\%) of having high carbohydrate days and regular meals days; 18.0\% of their dietary diary was found to be low carbohydrate days.

@@

Among the three types of carbohydrate eaters, the mean of total energy intake over the 4 days of dietary survey was the highest (7985.8 kJ, 95\%CI: 7283.3, 8146.3) in the low carbohydrate eaters group, and the lowest (7341.8 kJ, 95\%CI: 7172.5, 7511.2) in the moderate eaters group \textbf{(Table \ref{tab:tab-nutri-indi})}. Sources of energy for each type of carbohydrate eaters by the 7-time slots were also different. Low carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-A)} never had carbohydrate contributed more than 50\% of their total energy throughout the day. Energy from fat was the highest for low carbohydrate eaters most of the time during the day (except for the time slot between 10 pm to 6 am next morning). Most impressively, energy contributed by alcohol was always the highest in low carbohydrate eaters, percentages for energy from alcohol for the 7-times slots were 1.4\% (6-9 am), 1.8\% (9-12 noon), 3.1\% (12-2 pm), 7.2\% (2-5 pm), 8.4\% (5-8 pm), 20.0\% (8-10 pm), and 31.2\% (10-6 am), respectively. Contributions from different energy sources were quite similar for moderate and high carbohydrate eaters, but their absolute amount of energy consumption at each time slot was largely different. Moderate carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-B)} were characterised as consuming the lowest energy (296.9 kJ) before 9 am but having higher energy consumption (967.9 kJ) between 9 am and 12 noon time compared with low and high carbohydrate eaters. Moderate carbohydrate eaters may tend to have later breakfast, later lunch, and probably later dinner as well. They had the highest total energy consumption (599.7 kJ) at night (10 pm - 6 am) across all three types of eaters. High carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-C)} consumed the highest total energy (929.0 kJ) during 6 am to 9 am in the morning and the lowest total energy between 10 pm to 6 am (205.5 kJ). Specifically, carbohydrate contributions to total energy intake were 65.3\% (6-9 am), 56.2\% (9-12 noon), 46.2\% (12-2 pm), 50.3\% (2-5 pm), 42.5\% (5-8 pm), 44.1\% (8-10 pm), and 49.9\% (10-6 am). We also noticed that high carbohydrate eaters consumed their energy mainly from the following three time slots: 6-9 am, 12-2 pm, and 5-8 pm.

In total, the mean of total carbohydrate intake was 203.8 g, 218.3 g, and 233.4 g for low, moderate, and high carbohydrate eaters, respectively \textbf{(Table \ref{tab:tab-nutri-indi})}. Total energy contributed by carbohydrate was close to 50\% among the high carbohydrate eaters but was only 40.6\% among the low carbohydrate eaters. In terms of the subtypes (components) of the carbohydrate consumed at each time slot, high carbohydrate eaters consumed more than 2 times (compared to low carbohydrate eaters) and nearly 4 times (against moderate carbohydrate eaters) the amount of sugar (37.9g 95\% CI: 36.8, 39.2) and non-milk extrinsic sugar (i.e. free sugar, 11.1g 95\%CI: 10.7, 11.6) between 6-9 am. Moderate carbohydrate eaters had carbohydrate intakes that were more spread out throughout the day. They consumed more sugar and starch during 9-12 noon, 2-5 pm, 8-10 pm, and 10-6 am. Low carbohydrate eaters turned to have similar temporal pattern of consuming carbohydrates with the high carbohydrate eaters, but the absolute amount of fibre, sugar, free sugar, and starch consumed were usually lower than that in the high carbohydrate eaters except for time slots of 2-5 pm, and 10-6 am. Strong evidence (\textit{p} < 0.001) suggested that the

@@

mean of total fibre consumption for low, moderate, and high carbohydrate eaters were different: 13.7g (95\%CI: 13.4, 14.0), 12.5g (95\%CI: 12.1, 12.9), and 14.7g (95\%CI: 14.4, 14.9) with all 95\% CI being exclusive to each other. It is also noteworthy that low carbohydrate eaters consumed the highest average amount of protein (79.9 g, 17.2\% of total energy), fat (74.7g, 35.4\% of total energy), and alcohol (20.8 g, 6.8\% of total energy) as we have described for \textbf{Figure \ref{fig:energysourcesCB}}.

The social-demographic characteristics of the UK adults according to their individual level latent class membership are shown in \textbf{Table \ref{tab:Level2tab1}}. Moderate carbohydrate eaters were relatively younger (\textit{p} < 0.001). Gender distribution across the three types of carbohydrate eaters was fairly even (\textit{p} = 0.119). The distribution of the carbohydrate eater types appears to be changing with the year of survey. Low carbohydrate eaters represented 32.5\% of the population in the first year of the survey but later dropped to lower than 30\% (lowest in the third year, 22.6\%) until the most recent year. The proportion of high carbohydrate eaters increased from 41.2\% to the highest (50.6\%) in the second year, but then started to decline to 38.4\% in the 8th year of survey (\textit{p} = 0.015). There was no evidence of any difference in employment status across three types of carbohydrate eaters. However, strong evidence suggested that high carbohydrate eaters had the highest proportion (61.3\%) of living with partner (\textit{p} < 0.001); moderate carbohydrate eaters had the lowest average income (27180.8 \textsterling/year), the highest proportion of non-white population (20.5\%), and a lower education level (23.3\% with degree of higher education) compared with either low or high carbohydrate eaters.

@@

From the results of blood tests, 6.9\% of low carbohydrate eaters were found to be diabetic (diagnosed by A1C > 6.5\%), while the percentages of diabetes in the moderate and high carbohydrate eaters were 3.5\%, and 4.1\% (\textit{p} < 0.011), respectively. Although there was some evidence (\textit{p} = 0.027) that fasting blood glucose level may be slightly higher in non-diabetic low carbohydrate eaters, the geometric mean for A1C was probably lower in moderate carbohydrate eaters (5.43, 95\%CI: 5.39, 5.47). Total cholesterol, HDL, and LDL were all lower in the moderate carbohydrate eaters, while no evidence of any difference of TG was found across three types of carbohydrate eaters.

@@

		A1C (\%) & 5.47 (5.44, 5.51) & 5.43 (5.39, 5.47) & 5.50 (5.48, 5.53) & 0.010\\

@@

		\multicolumn{5}{l}{{\scriptsize BMI body mass index; WC, waist circumference; A1C, haemoglobin A1c; DM, diabetes mellitus; TC, total cholesterol, }}\\

@@

		\multicolumn{5}{l}{{\scriptsize Glucose, A1C, TC, HDL, LDL, TG, and physical activity were expressed in geometric means (95\% CI) because }} \\

@@

		\multicolumn{5}{l}{{\scriptsize Variables from the blood tests (glucose and A1C) were weighted by blood sample weights, the other variables were}}\\ 
		\multicolumn{5}{l}{{\scriptsize  weighted by nurse visiting weights. Glucose and A1C levels are estimated in subgroups of people without diabetes.}}\\

@@

		\multicolumn{5}{l}{{\scriptsize \textsuperscript{\dag} Hypertension was defined as either systolic blood pressure $\geqslant$ 140 mmHg or diastolic blood pressure $\geqslant$ 90 mmHg,}}\\
		\multicolumn{5}{l}{{\scriptsize or under treatment for hypertension.}}\\
%		\multicolumn{5}{l}{{\scriptsize \textsuperscript{\ddag} Carbohydrate percent indicates the percentage of energy from carbohydrate in total energy intake.}}\\  
		\multicolumn{5}{l}{{\scriptsize \textsuperscript{\S} DM was defined by A1C $>$ 6.5\%.}}\\
	\end{tabular}
\end{table}

@@

The sex-specific associations of carbohydrate eating patterns with hypertension (both in total and in participants without diabetes) are shown in \textbf{Table \ref{tab:tab2HYT}}. In the crude models, moderate carbohydrate eaters had statistically significant lower odds of having hypertension than low carbohydrate eaters in both men and women irrespective to diabetes status. Among men, after adjustment for selected confounders, which includes: age, live with partner or not, education level, BMI, smoking status, and total energy intake, the odds ratio (OR) comparing moderate with low carbohydrate eaters was 0.68 (95\% CI: 0.43, 1.07) and remained borderline significant (\textit{p} = 0.093). 95\% CI of the adjusted OR became narrower (OR: 0.64, 95\% CI: 0.41, 1.01, \textit{p} = 0.054) when BMI was replaced with WC in model 2. When diabetic men were excluded in Model 2, the odds for moderate and high carbohydrate eaters compared with low carbohydrate eaters were 35\% (OR: 0.65, 95\%CI: 0.41, 1.03) and 27\% (OR:0.73, 95\%CI: 0.51, 1.06) lower, respectively. The negative associations between moderate carbohydrate eating pattern and hypertension were also observed in women, however, without any statistically significant evidence in the fully adjusted models. High carbohydrate eaters also had lower adjusted odds compared with low carbohydrate eaters, while the 95\% CIs for the adjusted ORs were all wide and included the null value suggesting no evidence of any association in either men or women for high carbohydrate eating pattern and hypertension.

@@

		\multicolumn{7}{l}{{\scriptsize \textsuperscript{\S}  DM was defined by A1C $>$ 6.5\%.}}

@@

	\caption{\label{tab:tab2HYT} ORs (95\%CI) for the associations of carbohydrate eating patterns with hypertension in the UK adults, with or without diabetes .\\ (NDNS RP 2008/09-15/16, sample size = 6155)}\vspace{-0.3cm}
	\centering
	\fontsize{9}{11}\selectfont
	\begin{tabular}[t]{lccccc}
		\hiderowcolors
		\toprule
		\multicolumn{1}{c}{ } & \multicolumn{5}{c}{\textbf{Carbohydrate eating patterns}} \\
		\cmidrule(l{2pt}r{2pt}){2-6}
		\textbf{Model} & \textbf{Low} & \textbf{Moderate} & \textbf{\textit{P} value\textsuperscript{*}} & \textbf{High} & \textbf{\textit{P} value\textsuperscript{*}}\\
		\midrule
		\showrowcolors
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\textbf{Men (n = 2537)}}\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\hspace{1em}\textbf{Hypertension}}\\
		\hspace{1em}\hspace{1em}Crude model & 1 & 0.48 (0.33, 0.70) & < 0.001 & 0.82 (0.59, 1.13) & 0.217\\
		\hspace{1em}\hspace{1em}Model 1\textsuperscript{\dag} & 1 & 0.68 (0.43, 1.07) & 0.093 & 0.80 (0.56, 1.15) & 0.227\\
		\hspace{1em}\hspace{1em}Model 2 & 1 & 0.64 (0.41, 1.01) & 0.054 & 0.75 (0.53, 1.08) & 0.124\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\hspace{1em}\textbf{Hypertension in non-diabetics}}\\
		\hspace{1em}\hspace{1em}Crude model & 1 & 0.49 (0.33, 0.73) & < 0.001 & 0.82 (0.59, 1.14) & 0.241\\
		\hspace{1em}\hspace{1em}Model 1\textsuperscript{\dag} & 1 & 0.69 (0.43, 1.09) & 0.110 & 0.78 (0.54, 1.14) & 0.197\\
		\hspace{1em}\hspace{1em}Model 2 & 1 & 0.65 (0.41, 1.03) & 0.066 & 0.73 (0.51, 1.06) & 0.096\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\textbf{Women (n = 3618)}}\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\hspace{1em}\textbf{Hypertension}}\\
		\hspace{1em}\hspace{1em}Crude model & 1 & 0.52 (0.36, 0.75) & < 0.001 & 0.96 (0.72, 1.28) & 0.773\\
		\hspace{1em}\hspace{1em}Model 1\textsuperscript{\ddag} & 1 & 0.79 (0.45, 1.39) & 0.415 & 0.89 (0.61, 1.30) & 0.552\\
		\hspace{1em}\hspace{1em}Model 2 & 1 & 0.78 (0.45, 1.36) & 0.384 & 0.88 (0.62, 1.26) & 0.483\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\hspace{1em}\textbf{Hypertension in non-diabetics}}\\
		\hspace{1em}\hspace{1em}Crude model & 1 & 0.51 (0.35, 0.74) & < 0.001 & 0.98 (0.73, 1.31) & 0.875\\
		\hspace{1em}\hspace{1em}Model 1\textsuperscript{\ddag} & 1 & 0.79 (0.44, 1.42) & 0.435 & 0.89 (0.61, 1.29) & 0.534\\
		\hspace{1em}\hspace{1em}Model 2 & 1 & 0.79 (0.45, 1.39) & 0.415 & 0.87 (0.61, 1.25) & 0.452\\
		\bottomrule
		\multicolumn{6}{l}{{\scriptsize \textit{Note: }}}\\
		\multicolumn{6}{l}{{\scriptsize \textbf{Abbreviations}: OR, odds ratio; CI, confidence interval; BMI, body mass index; WC, waist}}\\
		\multicolumn{6}{l}{{\scriptsize  circumference; NDNS RP, national diet and nutrition survey rolling programme.}}\\
		\multicolumn{6}{l}{{\scriptsize Diabetes was defined by A1C > 6.5\%. BMI was replaced with WC in Model 2s, other}}\\
		\multicolumn{6}{l}{{\scriptsize covariates remained the same with the corresponding Model 1s.}}\\
		\multicolumn{6}{l}{{\scriptsize \textsuperscript{*} \textit{P} values were obtained from Wald tests in logistic regression models.}}\\ 
		\multicolumn{6}{l}{{\scriptsize \textsuperscript{\dag} Adjusted for age (continuous), live with partner or not (binary), education level}}\\ 
		\multicolumn{6}{l}{{\scriptsize (higher or equal to degree level or not), BMI, smoking status (current, ex-smoker, never),}}\\
		\multicolumn{6}{l}{{\scriptsize total energy intake (kJ);}}\\
		\multicolumn{6}{l}{{\scriptsize \textsuperscript{\ddag} Adjusted for age, live with partner or not, average household income (continuous),}}\\ 
		\multicolumn{6}{l}{{\scriptsize education level, BMI, smoking status, total energy intake (kJ), alcohol consumption (g/day).}}\\
	\end{tabular}
\end{table}
\vspace{-0.5cm}
\rowcolors{2}{white}{white}

@@

level (\textit{p} = 0.022 for men, < 0.001 for women), and average household income (\textit{p} = 0.011 for men, < 0.001 for women) decreased with increases in BMI. Living with a partner or not was strongly positively associated with obesity for men but not for women. Men with obesity were also found to have the lowest proportion of never being a smoker (47.9\%) and the highest proportion of being an ex-smoker (32.9\%). Association between smoking status and obesity in women had only very weak evidence (\textit{p} = 0.042), but a similar pattern as in men was observed (higher proportion of ex- and current smokers in overweight or obese women). No difference was found for length of physical activity across obesity levels in men, while in women, somewhat weak inverse association (\textit{p}=0.038) was confirmed. Interestingly, predefined carbohydrate eating patterns lacked evidence for an association with BMI in men (\textit{p} = 0.072) or in women (\textit{p} = 0.253).

@@

%			\multicolumn{12}{l}{{\scriptsize DM was defined by A1C > 6.5\%. BMI was replaced with WC in Model 2s, other covariates remained the same with the corresponding Model 1s.}}\\

>>>>

A series of traditional LCA of the responses to carbohydrate intake within the 7 time slots of the day was first examined. These initial analyses ignored the clustering of observation days within participants of the survey. \textbf{Table \ref{tab:mixmodels}} shows the latent class solutions for one to five classes (see rows under the Fixed effects model section). The BIC declines as the number of day level classes increases. However, the improvement of BIC dropped to less than 1000 from 3 class to 4 class solutions (658.9) and from 4 class to 5 class solutions (361.7). Entropy index indicates that the 4 class model could explain about 51\% percent of the data, while \textit{p} values of Lo-Mendell-Rubun LRT suggest that the more classes we fit, the better model we will have until up to 6 classes (\textit{p} = 0.06 and is not shown in the table). For the sake of parsimony, we only extended the model with random effects building on 2 class, 3 class, and 4 class solutions.

@@

As expected, consumption of total energy (7539.98 kJ), total carbohydrate (222.79 g), total sugar (98.63 g), starch (124.07 g), and non-milk extrinsic sugar (59.45 g) were the highest among high percentage carbohydrate days (all \textit{p} < 0.001). On the other hand, the consumption of protein (73.29 g), total fat (67.24 g), and alcohol (13.80 g) were the highest in the regular meals days. Moreover, in the high percentage carbohydrate days, participants turned out to consume the highest amount of fruit (107.40 g). There was no evidence of any difference in the consumption of yellow, red, or green vegetables across the three types of days (\textit{p} = 0.681).\vspace{-0.6cm}

@@

With three individual-level latent classes \textbf{(Figure \ref{fig:level2})}, a low-carbohydrate eaters class, a moderate-carbohydrate eaters class, and a high-carbohydrate eaters class emerge. 43.1\% participants were identified as high-carbohydrate eaters, in these individuals, about 50\% of the days (2 out of 4 days) of their dietary diary could be classified as having high percentage carbohydrate days. Nearly 1 out of 4 days of their dietary diary were either "regular meals day" or "low carbohydrate day". 28.1\% participants fell into the low carbohydrate eaters class on the left-hand side of \textbf{Figure \ref{fig:level2}}, their recordings of food intake showed that in more than 60\% of their days, they were having "regular meals" which was characterised as with the highest amount of fat and alcohol consumptions as already described in \textbf{Table \ref{tab:day-level-features}}. Moderate carbohydrate eaters have comparable proportions (42.0\% vs. 40.0\%) of high carbohydrate days and regular meals days; 18.0\% of their dietary diary was found to be low carbohydrate days.

@@

Among the three types of carbohydrate eaters, the mean of total energy intake over the 4 days of dietary survey was the highest (7985.8 kJ, 95\%CI: 7283.3, 8146.3) in the low carbohydrate eaters group, and the lowest (7341.8 kJ, 95\%CI: 7172.5, 7511.2) in the moderate eaters group \textbf{(Table \ref{tab:tab-nutri-indi})}. Sources of energy for each type of carbohydrate eaters by the 7-time slots were also different. Low carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-A)} never had carbohydrate contributing more than 50\% of their total energy throughout the day. Energy from fat was the highest for low carbohydrate eaters most of the time during the day (except for the time slot between 10 pm to 6 am next morning). Most impressively, energy contributed by alcohol was always the highest in low carbohydrate eaters, percentages for energy from alcohol for the 7-times slots were 1.4\% (6-9 am), 1.8\% (9-12 noon), 3.1\% (12-2 pm), 7.2\% (2-5 pm), 8.4\% (5-8 pm), 20.0\% (8-10 pm), and 31.2\% (10-6 am), respectively. Contributions from different energy sources were quite similar for moderate and high carbohydrate eaters, but their absolute amount of energy consumption at each time slot was largely different. Moderate carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-B)} were characterised as consuming the lowest energy (296.9 kJ) before 9 am but having higher energy consumption (967.9 kJ) between 9 am and 12 noon time compared with low and high carbohydrate eaters. Moderate carbohydrate eaters may tend to have later breakfast, later lunch, and probably later dinner as well. They had the highest total energy consumption (599.7 kJ) at night (10 pm - 6 am) across all three types of eaters. High carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-C)} consumed the highest total energy (929.0 kJ) during 6 am to 9 am in the morning and the lowest total energy between 10 pm to 6 am (205.5 kJ). Specifically, carbohydrate contributions to total energy intake were 65.3\% (6-9 am), 56.2\% (9-12 noon), 46.2\% (12-2 pm), 50.3\% (2-5 pm), 42.5\% (5-8 pm), 44.1\% (8-10 pm), and 49.9\% (10-6 am). We also noticed that high carbohydrate eaters consumed their energy mainly from the following three time slots: 6-9 am, 12-2 pm, and 5-8 pm.

In total, the mean of total carbohydrate intake was 203.8 g, 218.3 g, and 233.4 g for low, moderate, and high carbohydrate eaters, respectively \textbf{(Table \ref{tab:tab-nutri-indi})}. Total energy contributed by carbohydrate was close to 50\% among the high carbohydrate eaters but was only 40.6\% among the low carbohydrate eaters. In terms of the subtypes (components) of the carbohydrate consumed at each time slot, high carbohydrate eaters consumed more than 2 times (compared to low carbohydrate eaters) and nearly 4 times (against moderate carbohydrate eaters) the amount of sugar (37.9g 95\% CI: 36.8, 39.2) and non-milk extrinsic sugar (i.e. free sugar, 11.1g 95\%CI: 10.7, 11.6) between 6-9 am. Moderate carbohydrate eaters had carbohydrate intakes that were more spread out throughout the day. They consumed more sugar and starch during 9-12 noon, 2-5 pm, 8-10 pm, and 10-6 am. Low carbohydrate eaters turned to have similar temporal pattern of consuming carbohydrates as the high carbohydrate eaters, but the absolute amount of fibre, sugar, free sugar, and starch consumed were usually lower than that in the high carbohydrate eaters except for time slots of 2-5 pm, and 10-6 am. Strong evidence (\textit{p} < 0.001) suggested that the

@@

mean of total fibre consumption for low, moderate, and high carbohydrate eaters were different: 13.7g (95\%CI: 13.4, 14.0), 12.5g (95\%CI: 12.1, 12.9), and 14.7g (95\%CI: 14.4, 14.9) with all 95\% CI being non-overlapping. It is also noteworthy that low carbohydrate eaters consumed the highest average amount of protein (79.9 g, 17.2\% of total energy), fat (74.7g, 35.4\% of total energy), and alcohol (20.8 g, 6.8\% of total energy) as we have described for \textbf{Figure \ref{fig:energysourcesCB}}.

The social-demographic characteristics of the UK adults according to their individual level latent class membership are shown in \textbf{Table \ref{tab:Level2tab1}}. Moderate carbohydrate eaters were relatively younger (\textit{p} < 0.001). Gender distribution across the three types of carbohydrate eaters was fairly even (\textit{p} = 0.119). The distribution of the carbohydrate eater types appears to be changing with the year of survey. Low carbohydrate eaters represented 32.5\% of the population in the first year of the survey but later dropped to lower than 30\% (minimum in the third year at 22.6\%) until the most recent year. The proportion of high carbohydrate eaters increased from 41.2\% to the highest (50.6\%) in the second year, but then started to decline to 38.4\% in the 8th year of survey (\textit{p} = 0.015). There was no evidence of any difference in employment status across three types of carbohydrate eaters. However, strong evidence suggested that high carbohydrate eaters had the highest proportion (61.3\%) of living with partner (\textit{p} < 0.001); moderate carbohydrate eaters had the lowest average income (27180.8 \textsterling/year), the highest proportion of non-white population (20.5\%), and a lower education level (23.3\% with degree of higher education) compared with either low or high carbohydrate eaters.

@@

From the results of blood tests, 6.9\% of low carbohydrate eaters were found to be diabetic (diagnosed by HbA1C > 6.5\%), while the percentages of diabetes in the moderate and high carbohydrate eaters were 3.5\%, and 4.1\% (\textit{p} < 0.011), respectively. Although there was some evidence (\textit{p} = 0.027) that fasting blood glucose level may be slightly higher in non-diabetic low carbohydrate eaters, the geometric mean for HbA1C was probably lower (\textit{p} = 0.010) in moderate carbohydrate eaters (5.43, 95\%CI: 5.39, 5.47). Total cholesterol, HDL, and LDL were all lower in the moderate carbohydrate eaters, while no evidence of any difference of TG was found across three types of carbohydrate eaters.

@@

		HbA1C (\%) & 5.47 (5.44, 5.51) & 5.43 (5.39, 5.47) & 5.50 (5.48, 5.53) & 0.010\\

@@

		\multicolumn{5}{l}{{\scriptsize BMI body mass index; WC, waist circumference; HbA1C, haemoglobin HbA1C; DM, diabetes mellitus; TC, total cholesterol, }}\\

@@

		\multicolumn{5}{l}{{\scriptsize Glucose, HbA1C, TC, HDL, LDL, TG, and physical activity were expressed in geometric means (95\% CI) because }} \\

@@

		\multicolumn{5}{l}{{\scriptsize Variables from the blood tests (glucose and HbA1C) were weighted by blood sample weights, the other variables were}}\\ 
		\multicolumn{5}{l}{{\scriptsize  weighted by nurse visiting weights. Glucose and HbA1C levels are estimated in subgroups of people without diabetes.}}\\

@@

		\multicolumn{5}{l}{{\scriptsize \textsuperscript{\dag} Hypertension was defined as either systolic blood pressure $\geqslant$ 140 mmHg or diastolic blood pressure $\geqslant$ 90 mmHg,}}\\
		\multicolumn{5}{l}{{\scriptsize or under treatment for hypertension.}}\\
%		\multicolumn{5}{l}{{\scriptsize \textsuperscript{\ddag} Carbohydrate percent indicates the percentage of energy from carbohydrate in total energy intake.}}\\  
		\multicolumn{5}{l}{{\scriptsize \textsuperscript{\S} DM was defined by HbA1C $>$ 6.5\%.}}\\
	\end{tabular}
\end{table}

@@

The sex-specific associations of carbohydrate eating patterns with hypertension (both in total and in participants without diabetes) are shown in \textbf{Table \ref{tab:tab2HYT}}. In the crude models, moderate carbohydrate eaters had statistically significant lower odds of having hypertension than low carbohydrate eaters in both men and women irrespective to diabetes status. Among men, after adjustment for selected confounders, which includes: age, live with partner or not, education level, BMI, smoking status, and total energy intake, the odds of having hypertension were 32\% lower (OR: 0.68, 95\% CI: 0.43, 1.07) in moderate carbohydrate eaters compared with low carbohydrate eaters, evidence for the OR different from 1 was borderline significant (\textit{p} = 0.093). 95\% CI of the adjusted OR became narrower (OR: 0.64, 95\% CI: 0.41, 1.01, \textit{p} = 0.054) when BMI was replaced with WC in model 2. When diabetic men were excluded in Model 2, the odds for moderate and high carbohydrate eaters compared with low carbohydrate eaters were 35\% (OR: 0.65, 95\%CI: 0.41, 1.03) and 27\% (OR:0.73, 95\%CI: 0.51, 1.06) lower, respectively. The negative associations between moderate carbohydrate eating pattern and hypertension were also observed in women, however, without any statistically significant evidence in the fully adjusted models. High carbohydrate eaters also had lower adjusted odds compared with low carbohydrate eaters, while the 95\% CIs for the adjusted ORs were all wide and included the null value suggesting no evidence of any association in either men or women for high carbohydrate eating pattern and hypertension.

@@

		\multicolumn{7}{l}{{\scriptsize \textsuperscript{\S}  DM was defined by HbA1C $>$ 6.5\%.}}

@@

	\caption{\label{tab:tab2HYT} ORs (95\%CI) for the associations of carbohydrate eating patterns with hypertension in the UK adults, with or without diabetes .\\ (NDNS RP 2008/09-15/16, sample size = 6155)}\vspace{-0.3cm}
	\centering
	\fontsize{9}{11}\selectfont
	\begin{tabular}[t]{lccccc}
		\hiderowcolors
		\toprule
		\multicolumn{1}{c}{ } & \multicolumn{5}{c}{\textbf{Carbohydrate eating patterns}} \\
		\cmidrule(l{2pt}r{2pt}){2-6}
		\textbf{Model} & \textbf{Low} & \textbf{Moderate} & \textbf{\textit{P} value\textsuperscript{*}} & \textbf{High} & \textbf{\textit{P} value\textsuperscript{*}}\\
		\midrule
		\showrowcolors
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\textbf{Men (n = 2537)}}\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\hspace{1em}\textbf{Hypertension}}\\
		\hspace{1em}\hspace{1em}Crude model & 1 & 0.48 (0.33, 0.70) & < 0.001 & 0.82 (0.59, 1.13) & 0.217\\
		\hspace{1em}\hspace{1em}Model 1\textsuperscript{\dag} & 1 & 0.68 (0.43, 1.07) & 0.093 & 0.80 (0.56, 1.15) & 0.227\\
		\hspace{1em}\hspace{1em}Model 2 & 1 & 0.64 (0.41, 1.01) & 0.054 & 0.75 (0.53, 1.08) & 0.124\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\hspace{1em}\textbf{Hypertension in non-diabetics}}\\
		\hspace{1em}\hspace{1em}Crude model & 1 & 0.49 (0.33, 0.73) & < 0.001 & 0.82 (0.59, 1.14) & 0.241\\
		\hspace{1em}\hspace{1em}Model 1\textsuperscript{\dag} & 1 & 0.69 (0.43, 1.09) & 0.110 & 0.78 (0.54, 1.14) & 0.197\\
		\hspace{1em}\hspace{1em}Model 2 & 1 & 0.65 (0.41, 1.03) & 0.066 & 0.73 (0.51, 1.06) & 0.096\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\textbf{Women (n = 3618)}}\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\hspace{1em}\textbf{Hypertension}}\\
		\hspace{1em}\hspace{1em}Crude model & 1 & 0.52 (0.36, 0.75) & < 0.001 & 0.96 (0.72, 1.28) & 0.773\\
		\hspace{1em}\hspace{1em}Model 1\textsuperscript{\ddag} & 1 & 0.79 (0.45, 1.39) & 0.415 & 0.89 (0.61, 1.30) & 0.552\\
		\hspace{1em}\hspace{1em}Model 2 & 1 & 0.78 (0.45, 1.36) & 0.384 & 0.88 (0.62, 1.26) & 0.483\\
		\addlinespace[0.3em]
		\multicolumn{6}{l}{\hspace{1em}\textbf{Hypertension in non-diabetics}}\\
		\hspace{1em}\hspace{1em}Crude model & 1 & 0.51 (0.35, 0.74) & < 0.001 & 0.98 (0.73, 1.31) & 0.875\\
		\hspace{1em}\hspace{1em}Model 1\textsuperscript{\ddag} & 1 & 0.79 (0.44, 1.42) & 0.435 & 0.89 (0.61, 1.29) & 0.534\\
		\hspace{1em}\hspace{1em}Model 2 & 1 & 0.79 (0.45, 1.39) & 0.415 & 0.87 (0.61, 1.25) & 0.452\\
		\bottomrule
		\multicolumn{6}{l}{{\scriptsize \textit{Note: }}}\\
		\multicolumn{6}{l}{{\scriptsize \textbf{Abbreviations}: OR, odds ratio; CI, confidence interval; BMI, body mass index; WC, waist}}\\
		\multicolumn{6}{l}{{\scriptsize  circumference; NDNS RP, national diet and nutrition survey rolling programme.}}\\
		\multicolumn{6}{l}{{\scriptsize Diabetes was defined by HbA1C > 6.5\%. BMI was replaced with WC in Model 2s, other}}\\
		\multicolumn{6}{l}{{\scriptsize covariates remained the same with the corresponding Model 1s.}}\\
		\multicolumn{6}{l}{{\scriptsize \textsuperscript{*} \textit{P} values were obtained from Wald tests in logistic regression models.}}\\ 
		\multicolumn{6}{l}{{\scriptsize \textsuperscript{\dag} Adjusted for age (continuous), live with partner or not (binary), education level}}\\ 
		\multicolumn{6}{l}{{\scriptsize (higher or equal to degree level or not), BMI, smoking status (current, ex-smoker, never),}}\\
		\multicolumn{6}{l}{{\scriptsize total energy intake (kJ);}}\\
		\multicolumn{6}{l}{{\scriptsize \textsuperscript{\ddag} Adjusted for age, live with partner or not, average household income (continuous),}}\\ 
		\multicolumn{6}{l}{{\scriptsize education level, BMI, smoking status, total energy intake (kJ), alcohol consumption (g/day).}}\\
	\end{tabular}
\end{table}
\vspace{-0.5cm}
\rowcolors{2}{white}{white}

@@

level (\textit{p} = 0.022 for men, < 0.001 for women), and average household income (\textit{p} = 0.011 for men, < 0.001 for women) decreased with increases in BMI. Living with a partner was strongly positively associated with obesity for men but not for women. Men with obesity were also found to have the lowest proportion of never being a smoker (47.9\%) and the highest proportion of being an ex-smoker (32.9\%). Association between smoking status and obesity in women had only very weak evidence (\textit{p} = 0.042), but a similar pattern as in men was observed (higher proportion of ex- and current smokers in overweight or obese women). No difference was found for length of physical activity across obesity levels in men, while in women, somewhat weak inverse association (\textit{p}=0.038) was confirmed. Interestingly, predefined carbohydrate eating patterns lacked evidence for an association with BMI in men (\textit{p} = 0.072) or in women (\textit{p} = 0.253).

@@

%			\multicolumn{12}{l}{{\scriptsize DM was defined by HbA1C > 6.5\%. BMI was replaced with WC in Model 2s, other covariates remained the same with the corresponding Model 1s.}}\\
